FFT based convolution

### 2010-03-06 article/Article.tex
article continuated
Ignore-this: f0d0b161df09c007b5e4da7133fcf2a1

darcs-hash:20100306233314-2fc9d-b3a8c2d79931cf5617e6916014410ccc2e80947f.gz

\subsection{FFTW filters enhancements}
\subsubsection{Thread support}
fft et norm
\subsubsection{Memory usage reduction}
\subsubsection{Code safety}

>>>>

\subsection{FFTW filters enhancements}

ITK provides FFT and inverse FFT filters based on the FFTW library. Those filters had some problems,
or were missing some features. They have been enhanced to fit the needs of the FFT based convolution
filter.

\subsubsection{Thread support}
The FFTW filters were not multithreaded. There are two places were multithreading can take place: the
FFTW execution and the normalization step.

The FFTW execution can be threaded simply by calling $fftwf_plan_with_nthreads(threads);$ or $fftwf_plan_with_nthreads(threads);$, depending on the pixel type, before creating the plan. 
itk::fftw::Proxy has been modified to provide this feature: an extra $thread$ parameter, which default
to one to keep the backward compatibility, has been added to the method in charge of creating the plans.
Then the itk::FFTWComplexConjugateToRealImageFilter and itk::FFTWRealToComplexConjugateImageFilter have
been modified to pass their user defined number of threads to those methods.

In order to implement the threading support in the normalization step in 
itk::FFTWComplexConjugateToRealImageFilter, the execution of the FFTW library has been moved to 
BeforeThreadedGenerateData(), and the normalization step to ThreadedGenerateData().

\subsubsection{Memory usage reduction}

FFTW filters were using a lot of memory, because they were allocating an internal buffer for the input 
image, and an other one for the output image. Worth, these buffer were not deallocated at the end of
the execution filter, keeping the memory usage very high even when the execution has completed.

I guess the main reasons to do that were:
\begin{itemize}
  \item make sure that the buffers are properly aligned.
  \item avoid the destruction of the input. The creation of the FFTW plan, and the execution of FFTW 
  can destroy the input. This is most of the time not acceptable in ITK.
  \item reuse the FFTW for a latter run. The plan include the memory location of the input and output
  buffer, so the buffer can't be deallocated, as it may be the case with the filter input and output images.
\end{itemize}

The buffer were simply allocated with the $new$ operator, which is not enough to ensure a proper alignement
on some plateforms, including Microsoft Windows and Linux.

The destruction of the input can be avoided in the creation plan by using the $FFTW_ESTIMATE$ flag
-- that flag was already used in the implementation. It can also be avoided while running FFTW, for 
the real to complex transform, by also using the flag $FFTW_PRESERVE_INPUT$, so in 
itk::FFTWRealToComplexConjugateImageFilter, there is no need for the input buffer. The case is a bit
more difficult for the complex to real transform: FFTW does not provide any algorithm able to preserve
the input, so the input must be copied to an internal buffer. There is a case however, when we don't
care about destroying the input in ITK: when the ReleaseDataFlag is ON on the input. In that case, 
itk::FFTWRealToComplexConjugateImageFilter has been modified to use the input image directly without 
intermediate buffer. The detection of the ReleaseDataFlag is a bit difficult during the pipeline 
execution: this flag is always modified to OFF before the call to $GenerateData()$ to avoid the 
destruction by a minipipeline. In consequence, the detection of the ReleaseDataFlag is done at an 
earlier stage in the pipeline execution, in $UpdateOutputData()$, and stored in a member variable.

The latter reuse of the plan, while sensible when not using the $FFTW_ESTIMATE$ flag, is not much 
interesting when it is used: the time needed to create the plan is very short -- way shorter than the
FFT computation. Also, the number of threads and the image size are stored in the plan, and can't be 
changed later, which doesn't fit well with ITK pipeline model. In the enhanced version, the plan is
not kept between the filters execution, and so no intermediate output buffer is used, decreasing 
significantly the needed memory.

\subsubsection{Code safety}

Only $fftwf_execute(plan);$ and $fftw_execute(plan);$ are thread safe in FFTW -- the other functions
are not thread safe. In consequence, the call to all the other functions has been protected by a 
global lock in the enhanced classes.

Also, the check for the size of the image to know if the plan can be reused may not have been good 
enough, because it was testing only the number of pixels and not the size on all the dimensions. 
The plan is not kept between the execution, so this can be a problem anymore.
